\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-04-27 doc/publications/plos16/nm.tex
changes to  nm.tex

% Please keep the abstract below 300 words
\section{Abstract}

Analyses are being performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
they scale well with samples ($n$) but not with features ($p$, variants).

@@

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the health space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

Identifying the relevant or disease specific base pair differences between individuals is the focus of genome wide
association studies (GWAS).  These analyses are typically performed on only the most frequently differing base pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011} and the approach should 
in-fact be extended to interrogating all genomic positions[CITATION WES vs WGS?].

Hence using more sophisticated machine learning (ML) approaches, in particular tree-based models, have been successful
for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit data sets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelise. As the forest is a sum of decision trees, it is possible to grow separate
trees on different processors and combine the trees. 
%See the discussion of the properties of random forests in Section \ref{section:methods}.

@@

$$
\ddot{y} =    \left\{ \begin{array}{rcl}
			  0 & \mbox{for} & y \leq Q2(y) \\
			  1 & \mbox{for} & y > Q2(y).
		 \end{array}\right 
$$

>>>>

% Please keep the abstract below 300 words
\section{Abstract}
Modern data analyses are performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
they scale well with samples ($n$) but not with features ($p$, variants).

@@

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the biomedical space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

Identifying the relevant or disease specific genetic variations between individuals is the focus of genome wide
association studies (GWAS).  These analyses are typically performed on only the most frequently differing base pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011} and the approach should 
in-fact be extended to interrogating all genomic positions[CITATION WES vs WGS?].

Hence using more sophisticated machine learning (ML) approaches, in particular tree-based models, have been successful
for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests \cite{Breiman2001} are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit data sets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelise. As the forest is a sum of decision trees, it is possible to grow separate
trees on different compute nodes and combine the trees. 
%See the discussion of the properties of random forests in Section \ref{section:methods}.

@@

$$
\ddot{y} =    \left\{ \begin{array}{rcl}
			  0 & \mbox{for} & y \leq Q2(y) \\
			  1 & \mbox{for} & y > Q2(y).
		 \end{array}\right. 
$$
